---
title: "Alpine Swift Migration Dashboard - Technical Report"
subtitle: "CCES FS26 - Interactive Environmental Data Dashboard"
author:
  - name: "Lukas Buchmann"
    affiliation: "ZHAW - Environmental Science"
  - name: "Etienne Duerig"
    affiliation: "ZHAW - Environmental Science"
date: today
abstract: |
  This report documents an interactive R Shiny dashboard that visualises
  the trans-Saharan migration of the Alpine Swift (*Tachymarptis melba*).
  Light-level geolocator tracks for up to 215 individuals from nine
  colonies across four populations are loaded directly from the Movebank
  Data Repository [@movebank]. A single integrated map combines an
  always-on Path-Resample dot cloud with a server-driven day-of-year
  animation; the inherent latitudinal uncertainty of light-level
  geolocators is rendered explicitly as a shadow band around each
  migration fix. Cartographic and visual-analytic decisions follow
  @slocum2009, @maceachren1995, @maceachren2004, @cairo2016 and
  @keim2008. The intended audience is non-specialist birding
  enthusiasts.
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
    fig-cap-location: bottom
    theme: cosmo
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    fig-cap-location: bottom
    geometry:
      - margin=2.5cm
    documentclass: article
execute:
  echo: true
  warning: false
  message: false
  cache: false
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
suppressPackageStartupMessages({
  library(dplyr); library(ggplot2); library(readr); library(sf)
  library(lubridate); library(scales); library(viridisLite)
})
COUNTRY_COLORS <- c(Switzerland = "#1B9E77", Spain = "#D95F02",
                    Bulgaria   = "#7570B3", Turkey = "#E7298A")
```

# Dashboard context and goals

The Alpine Swift (*Tachymarptis melba*, formerly *Apus melba*) is a small aerial insectivore capable of one of the longest non-stop flights known among birds; individuals can remain airborne for up to 200 consecutive days outside the breeding season, feeding, sleeping and moulting on the wing [@meier2020]. Conventional ring recoveries say almost nothing about where these birds go between September and March, yet conserving aerial insectivores under European climate change requires precisely that knowledge.

Light-level geolocators have made the trans-Saharan annual cycle tractable for the first time. Since 2014, the Swiss Ornithological Institute has tagged Alpine Swifts at four populations along a twelve-degree latitudinal gradient: Pirasali Island (Turkey, 36.3 deg N), Tarragona (Spain, 41.1 deg N), Sofia (Bulgaria, 42.7 deg N) and six colonies in Switzerland (46-47 deg N). The headline finding of @meier2020 is that the eastern and western populations follow **two geographically distinct flyways** across the Sahara - the western pair via Iberia and Western Sahara, the eastern pair via the eastern Mediterranean and Egypt - and winter in **spatially separate** regions of tropical Africa, even though their migration speeds are statistically indistinguishable. Length of stay at the breeding site is negatively correlated with breeding latitude.

The dashboard turns that scientific result into an interactive experience for **bird enthusiasts**, with three goals: (i) make the spatial separation of the two flyways visually obvious, (ii) let users explore how breeding latitude shapes annual-cycle phenology, and (iii) communicate the rapidity of the trans-Saharan crossing through a playable animation in which geolocator uncertainty is shown rather than hidden.

# MacEachren's Geovisualization Cube and our audience

[@maceachren1995; @maceachren2004] places any cartographic interface inside a three-dimensional design space - the *Geovisualization Cube*. The three axes are *audience* (private to public), *task* (exploring to presenting) and *interaction* (low to high). Conventional thematic maps in a printed journal sit at the *public-presenting-low* corner; internal exploratory scripts sit at the opposite *private-exploring-high* corner.

The Alpine Swift dashboard deliberately occupies a **middle position**: *public, reveal/present, medium-to-high interaction*. The intended audience is bird enthusiasts who already love the species. That choice constrains every later decision:

- The dashboard explains itself - About tab, adaptive on-map legend and per-layer hover labels substitute for an analyst's prior knowledge.
- Defaults are narratively useful, not statistically neutral. The slider opens at day 100 (early April), when Alpine Swifts are at the breeding sites, so the first frame is recognisable.
- The interaction repertoire is broad but bounded: filter, brush-via-click, animate and overlay-toggle, but not raw-data export or statistical-model controls - those belong to the private-exploring corner of the cube.

# The Cairo Wheel evaluation

@cairo2016 proposes a six-axis Visualisation Wheel; each axis is a trade-off, and a good visualisation is one whose chosen positions are internally consistent with the intended audience, not "as far to one side as possible".

| Cairo axis | Our position | Justification |
|------------------------|------------------------|------------------------|
| Abstraction - Figuration | Mid, toward abstraction | A subdued Positron basemap is figurative enough to be recognisable, suppressed enough that the bird tracks dominate. |
| Functionality - Decoration | Closer to functional | No chartjunk, no rainbow ramps. The only "decoration" is the warm KPI colouring matched to the lay audience. |
| Density - Lightness | Mid, adjustable | Path resamples are alpha-blended so density encodes credibility; uncertainty and trail layers are individually toggleable. |
| Multidimensionality - Unidimensionality | Multidimensional | Population, time, phase, latitude and uncertainty are encoded simultaneously because flyway separation is a spatio-temporal phenomenon. |
| Originality - Familiarity | Conventional with one bespoke layer | Boxplots, time sliders and Leaflet basemaps are conventional. The migration-uncertainty shadow is the one bespoke device, derived from Figure 1 of @meier2020. |
| Novelty - Redundancy | Slightly redundant | The current day-of-year is shown in three places (map highlight, slider thumb, date readout); Cairo treats this purposeful redundancy as a feature, not a flaw. |

# The Visual Analytics mantra

@keim2008 condenses Visual Analytics into "*Analyse first - show the important - zoom, filter and analyse further - details on demand*", an extension of @shneiderman1996's earlier "*Overview first, zoom and filter, details on demand*". The dashboard implements all four steps:

1.  **Analyse first.** `R/data_processing.R` does the heavy work before any pixel is drawn: per-bird airspeed filtering (drops fixes that would require \> 50 km/h ground speed, just above the Meier 2020 cruising speed of 12.6 m/s = 45.4 km/h), daily median aggregation, and the rolling 10/90 percentile latitudinal uncertainty band.
2.  **Show the important.** The first frame already shows the two flyways under the default *Color by Country* setting - the headline finding of @meier2020 is the opening view.
3.  **Zoom, filter, analyse further.** Sidebar quick filters (All / W flyway / E flyway / Clear), the per-colony picker, the year slider, the layer-control overlay toggles and Leaflet's native pan-zoom give the user explicit control over spatial, temporal and categorical extent.
4.  **Details on demand.** Every dot carries a hover label and a click handler. Clicking selects the bird and draws its full bird-year tracks; the sidebar card shows colony, country, flyway, fix count, latitude range and observed phases. Clicking the same bird again deselects.

# Key geovisual analytics interactions

Following the @maceachren2004 taxonomy, the dashboard supports six classical interactions:

- **Identify** - hover labels and the on-map legend identify each dot's bird, colony and phase.
- **Locate** - map projection, layer control and animation date readout locate every observation in space and time.
- **Filter** - quick-filter buttons, the colony picker and the year slider restrict the active row set. The *Color by* radio is deliberately not a filter - it only re-codes existing dots, so the visible point set never changes silently.
- **Re-express / re-encode** - *Color by* Country / Colony / Flyway / Year re-expresses the same observations under different categorical decompositions. The on-map legend adapts in real time.
- **Compare** - the latitude boxplot beneath the map compares the current-day distribution across populations; the phenology card compares whole annual-cycle trajectories.
- **Animate / playback** - Play / Pause / Reset, three speed steps and the slider give the user complete time control.

# Geolocator uncertainty and how we visualise it

Light-level geolocators infer position from sunrise / sunset times. **Longitude** comes from the *time* of solar noon and is fairly accurate (\~30-100 km). **Latitude** comes from *day length* and is far less accurate (\~150 km nominal, growing to \~300 km near the equinoxes when day length is identical at all latitudes) [@lisovski2012; @lisovski2020]. Alpine Swifts breed in dark crevices, so cavity behaviour contaminates dawn / dusk estimates - the published data therefore omits localisation during the breeding phase. Wintering positions appear visibly grid-binned because the published FLightR posterior collapses to a discrete spatial grid during long stationary periods.

The dashboard surfaces this uncertainty explicitly rather than hiding it:

1.  **Path-Resample dot cloud.** Every retained daily fix is drawn simultaneously at low alpha; overlapping fixes accumulate into denser regions, so dot density itself encodes credibility (Meier 2020, Figure 1 style).
2.  **Migration-uncertainty shadow.** For every migration fix we draw a thin vertical polyline from `(lon, lat_lo)` to `(lon, lat_hi)`, with `lat_lo` / `lat_hi` the 10th and 90th percentiles of a centred 7-day rolling window per bird-year. The polyline is rendered at 10% opacity in the bird's colour, so cumulatively the shadows form a diffuse latitudinal band rather than a misleading hard line. Crucially, this shadow is *length-encoded* in latitude and *spatially anchored* to the fix's longitude, matching the underlying error structure; longitudinal error is much smaller and deliberately not visualised.
3.  **Burst-aware lines.** Phenology lines are split where the geolocator missed more than 14 consecutive days, so no straight line bridges a multi-week gap.
4.  **Airspeed filter.** Consecutive pairs that would require \> 50 km/h average ground speed are removed before aggregation, eliminating large light-noise spikes.

The shadow is toggleable in the map's layer control, so users who prefer a clean static frame can switch it off without losing the underlying dots.

# Visual variables, measurement levels and colour scheme

@stevens1946's measurement-level taxonomy and @bertin1983's semiology of graphics jointly motivate every channel decision in the dashboard. Nominal categories map to *hue*, ordinal categories to *lightness*, and ratio quantities to *length* or *position* - the Bertin / Slocum default, chosen because the human visual system is best at matching those visual variables to those measurement levels.

| Variable | Level | Visual variable | Implementation |
|------------------|------------------|------------------|------------------|
| `country` | Nominal | Hue | Dark2 4-colour palette. |
| `colony_name` | Nominal | Hue (extended) | Dark2 ramp, stable per colony. |
| `flyway` (W / E) | Nominal | Hue | First two Dark2 hues - same family as `country` so legends remain consistent. |
| `year` (2014-2016) | Ordinal | Lightness | YlGnBu 3-step sequential ramp. |
| `phase` | Nominal | Alpha | Migration dots 0.55, breeding / wintering 0.28. |
| `lat` | Ratio | Position (y) | Latitude has a meaningful zero at the equator. |
| `lon` | Interval | Position (x) | Longitude has an arbitrary zero (Greenwich). |
| `doy` (day of year) | Interval (cyclical) | Slider position + animation tick |  |
| `lat_unc = lat_hi - lat_lo` | Ratio | Length of shadow stroke | Doubled length = doubled uncertainty. |
| `n_birds_active_on_day` | Ratio | Boxplot whisker + jitter density | Zero is meaningful. |

Qualitative groupings use **ColorBrewer "Dark2"** (Switzerland #1B9E77, Spain #D95F02, Bulgaria #7570B3, Turkey #E7298A): canonical for nominal data with up to eight categories, colour-blind safe under protanopia, deuteranopia and tritanopia, and reliably distinguishable on print. The flyway palette reuses the first two Dark2 hues so the legend stays consistent between *Color by Country* and *Color by Flyway*. Ordinal `year` uses **YlGnBu** (#EDF8B1 -\> #7FCDBB -\> #2C7FB8), a multi-hue sequential ramp that survives black-and-white printing and reads correctly under common forms of colour blindness. **No rainbow / spectral ramps are used anywhere.** The basemap is desaturated CartoDB Positron (figure / ground principle). Selected-bird highlights use black + sodium yellow (#FFE45E), a chromatic family not used by any population palette, so the selection layer never collides with the group colours.

For the animation, the same channel discipline carries through: hue continues to encode population identity across frames; symbol size and a dark outline lift the current frame above the static cloud without changing hue; an explicit Leaflet pane (`paneCurrent`, z-index 470) guarantees that the current-day markers always sit on top; phase-aware alpha keeps the breeding cluster from saturating the map.

# Additional points: MAUP and the moving-point trail

**MAUP.** The Modifiable Areal Unit Problem [@openshaw1984] arises when point observations are summarised inside areal units (countries, hex grids) and the choice of boundary materially changes the conclusion. The dashboard mostly avoids MAUP because it visualises *individual point fixes*, not area-aggregated counts. Two latent sources remain. The per-bird phase labels (breeding / migration / wintering) come from a hand-curated classification in the Movebank `comments` column - the boundaries of those classes are themselves a temporal areal decision. We mitigate this by keeping the original authors' labels and by showing the lat_lo / lat_hi shadow during migration so that spatial uncertainty is not silenced by the phase label. The FLightR grid-binned wintering positions are a second latent source; we deliberately do not re-aggregate them so that the user sees the grid artefact explicitly rather than smoothed into a misleadingly continuous distribution.

**Why a moving-point trail (sperm trail).** A naive animation that only redraws the current frame loses recent direction and speed - precisely the property that makes Alpine Swifts remarkable (median trans-Saharan crossing 6 days outbound, 9 days return). The toggleable Moving-point Trail draws each bird's last 1 / 3 / 7 days as a continuous polyline coloured by the active grouping. Short trails communicate apparent motion via Wertheimer's *phi phenomenon* without forcing a high frame rate. Because we never interpolate across gaps, the trail visibly *breaks* where the bird was untracked - honest about the gap rather than silently filling it. The four-level radio (Off / 1 / 3 / 7 days) lets the user choose how much past context they want.

# Static figures

```{r}
#| label: fig-flyways
#| fig-cap: "Path-resample dot cloud for the trans-Saharan annual cycle, reprojected to an equal-area projection (Eckert IV, EPSG:54012). Each translucent dot is one retained daily fix; the two distinct flyways (western: Switzerland + Spain; eastern: Bulgaria + Turkey) and their spatially separate wintering grounds at 5-10 deg N are visible as denser regions of the cloud."
#| fig-width: 8
#| fig-height: 5

mb_dir <- "data/raw/movebank"
files  <- list.files(mb_dir, pattern = "tracks.*\\.csv$",
                     full.names = TRUE, ignore.case = TRUE)

if (length(files) == 0 && file.exists("data/raw/tracks.csv")) {
  tr <- readr::read_csv("data/raw/tracks.csv", show_col_types = FALSE)
} else if (length(files) > 0) {
  tr <- purrr::map_df(files, function(f) {
    raw  <- readr::read_csv(f, show_col_types = FALSE, progress = FALSE)
    nm   <- names(raw)
    lonc <- nm[grep("location.{0,1}long|longitude$", nm, ignore.case = TRUE)][1]
    latc <- nm[grep("location.{0,1}lat|latitude$",  nm, ignore.case = TRUE)][1]
    sn   <- basename(f)
    country <- dplyr::case_when(
      grepl("Switzerland", sn) ~ "Switzerland",
      grepl("Spain",       sn) ~ "Spain",
      grepl("Bulgaria",    sn) ~ "Bulgaria",
      grepl("Turkey",      sn) ~ "Turkey",
      TRUE                     ~ NA_character_)
    data.frame(lon = raw[[lonc]], lat = raw[[latc]], country = country)
  })
}

set.seed(2026)
tr <- tr[!is.na(tr$lat) & !is.na(tr$lon) & !is.na(tr$country), ]
if (nrow(tr) > 12000) tr <- tr[sample(nrow(tr), 12000), ]

sf_pts <- st_as_sf(tr, coords = c("lon", "lat"), crs = 4326) |>
  st_transform("ESRI:54012")

ggplot(sf_pts) +
  geom_sf(aes(color = country), size = 0.4, alpha = 0.30) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey80") +
  scale_color_manual(values = COUNTRY_COLORS, name = NULL) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid = element_line(color = "grey92"))
```

```{r}
#| label: fig-phenology
#| fig-cap: "Latitude across the annual cycle for every bird-year tracked in the Movebank Data Repository, coloured by country. Each thin line is one continuous tracking segment of a single bird-year; lines are split where the geolocator missed more than 14 consecutive days, so the summer breeding gap (May - August, when birds are inside dark nest cavities and the device cannot estimate position) stays visibly empty instead of being bridged by fake interpolation. Steep descents in September - October are autumn migration; the rebounds in March - April are the spring return. All populations winter at 5 - 10 deg N. Higher-latitude populations (Switzerland, Bulgaria) show shorter breeding-site occupancy than lower-latitude ones (Spain, Turkey) - the latitudinal-compression result of @meier2020."
#| fig-width: 8
#| fig-height: 3.6

if (file.exists("data/processed/tracks_processed.rds")) {
  pr <- readRDS("data/processed/tracks_processed.rds")
  d  <- pr$daily |>
    dplyr::arrange(bird_id, year, date) |>
    dplyr::group_by(bird_id, year) |>
    dplyr::mutate(
      days_diff   = as.numeric(difftime(date, dplyr::lag(date),
                                        units = "days")),
      new_segment = ifelse(is.na(days_diff) | days_diff > 14, 1, 0),
      segment_id  = cumsum(new_segment)
    ) |>
    dplyr::ungroup() |>
    dplyr::mutate(line_group = paste(bird_id, year, segment_id))

  ggplot(d, aes(x = doy, y = lat, color = country, group = line_group)) +
    geom_line(alpha = 0.22, linewidth = 0.35) +
    scale_color_manual(values = COUNTRY_COLORS, name = NULL) +
    scale_x_continuous(
      breaks = c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335),
      labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                 "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
      expand = expansion(mult = c(0.005, 0.005))
    ) +
    scale_y_continuous(limits = c(0, 50),
                       breaks = seq(0, 50, by = 10)) +
    labs(x = NULL, y = "Latitude (degrees N)") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank())
}
```

# Implementation notes

- **Modular Shiny.** Each panel is a self-contained module (`mod_*.R`) using `moduleServer()`. Cross-module communication flows through the `filters_server()` return value (a list of reactives), so dependencies are one-directional and the dashboard is testable.
- **Server-side animation.** The loop uses `reactiveVal(is_playing)`, `invalidateLater(200)` and `updateSliderInput()`. A throttled `plot_doy()` reactive shields the boxplot from per-tick re-renders.
- **Leaflet panes for honest z-ordering.** Five panes are declared at map creation (`paneShadow` 380, `paneTrail` 410, `paneResamples` 425, `paneSelected` 440, `paneCurrent` 470). Every layer is tagged with the appropriate pane, so the current-day markers are guaranteed to render on top irrespective of observer firing order.
- **Adaptive on-map legend.** A dedicated observer maintains a single controllable legend at the top-right using `removeControl("anim_legend")` + `addControl(..., layerId = "anim_legend")`. The legend HTML is generated by `render_map_legend(resolve_palette(df, gm))`, so it always matches the drawn data.
- **Cache.** The processed table (daily medians, uncertainty band, phenology) is persisted in `data/processed/tracks_processed.rds`; re-launches start in \~1 s. Rebuild with `build_processed_data(force = TRUE)`.
- **Portability.** Every path is relative; the project drops into any Positron / RStudio workspace without modification. A bundled synthetic dataset means the dashboard runs offline immediately when no Movebank files are present.

# Cartographic and scientific compliance

- @slocum2009**.** Figure / ground via a low-chroma Positron basemap (ch. 12), ColorBrewer Dark2 for nominal data and YlGnBu sequential for ordinal `year` (ch. 14), integrated rather than two-panel static / animated map (ch. 22), and an equal-area Eckert IV reprojection for the static report figures.
- @meier2020**.** The dashboard reads the four-population gradient, the published `comments` phase classification and the airspeed cap directly from the Meier methodology. The Path-Resample dot cloud and the migration uncertainty shadow are the interactive counterparts of the paper's Figure 1. No fake interpolation crosses gaps larger than 14 days.

# Limitations and future work

- **Wind support.** Pairing the tracks with ECMWF ERA-5 reanalysis would let users explore whether deviations from the corridor centroid correlate with daily wind anomalies - the mechanism the paper hypothesises for the east-west flyway divergence.
- **Kernel-density wintering surface.** A toggleable 2D kernel density (with bandwidth justified by geolocator uncertainty) could sharpen the wintering distinction without re-introducing a hull boundary.
- **Live Movebank pull.** A `move2`-driven option to refresh the cache from the Movebank API on demand would make the dashboard truly live.

# Declarations

**Individual contributions.** Lukas Buchmann implemented the dashboard and reviewed the report draft. Etienne Duerig implemented the data pipeline and wrote the report.

**AI use.** Code scaffolding, the synthetic-data generator, polish iterations and a first pass of this report were produced with the assistance of an Anthropic Claude model. All generated code was reviewed line-by-line before commit; the cartographic choices, the storytelling structure and the selection of the Phase-2 extension are the authors' own. No text from @meier2020 is reproduced verbatim.

# References {.unnumbered}

::: {#refs}
:::