---
title: "Alpine Swift Migration Dashboard — Technical Report"
subtitle: "CCES FS26 · Interactive Environmental Data Dashboard"
author:
  - name: "Lukas Buchmann"
    affiliation: "ZHAW · Environmental Science"
  - name: "Etienne Dürig"
    
date: today
abstract: |
  This report documents an interactive R Shiny dashboard that visualises
  the trans-Saharan migration of the Alpine Swift (*Tachymarptis melba*).
  Light-level geolocator tracks for 173 individuals from nine colonies
  across four populations are loaded directly from the Movebank Data
  Repository. Three coupled views — a Leaflet-based spatial map, two
  linked phenology plots, and an animated replay of the annual cycle —
  surface the headline finding of Meier et al. (2020): the eastern and
  western populations follow geographically distinct flyways across the
  Sahara and winter in spatially separate regions of tropical Africa,
  despite indistinguishable migration speeds. Cartographic decisions
  follow Slocum et al. (*Thematic Cartography and Geovisualization*):
  perceptually uniform sequential ramps, colour-blind-safe qualitative
  palettes, figure-ground hierarchy, and an area-faithful representation
  of the Europe-Africa extent in the static report figures.
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
    fig-cap-location: bottom
    theme: cosmo
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    fig-cap-location: bottom
    geometry:
      - margin=2.5cm
    documentclass: article
execute:
  echo: true
  warning: false
  message: false
  cache: false
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
suppressPackageStartupMessages({
  library(dplyr); library(ggplot2); library(readr); library(sf)
  library(lubridate); library(scales); library(viridisLite)
})
COUNTRY_COLORS <- c(Switzerland = "#1B9E77", Spain = "#D95F02",
                    Bulgaria   = "#7570B3", Turkey = "#E7298A")
```

# Dashboard context and goals

The Alpine Swift (*Tachymarptis melba*, formerly *Apus melba*) is a small
aerial insectivore that performs one of the longest non-stop flights
known among birds. The species can stay airborne for up to 200
consecutive days outside of the breeding season and is exclusively
aerial during that time [@meier2020]. Conventional bird-banding can
say almost nothing about where these birds *go* between September and
March — yet conserving aerial insectivores in a rapidly warming
Europe requires precisely that knowledge.

Light-level geolocators have made the trans-Saharan annual cycle of
European-breeding swifts tractable for the first time. The Swiss
Ornithological Institute has, since 2014, run a coordinated long-term
study tagging Alpine swifts at four breeding populations along a
roughly twelve-degree latitudinal gradient: Pırasalı Island
(Turkey, 36.3 °N), Tarragona (Spain, 41.1 °N), Sofia
(Bulgaria, 42.7 °N) and six colonies in Switzerland (46–47 °N).
The headline finding of @meier2020 is that the eastern and western
populations follow **two geographically distinct flyways** across the
Sahara — the western pair (Spain + Switzerland) crossing via the
Iberian Peninsula and Western Sahara, the eastern pair (Bulgaria +
Turkey) via the Eastern Mediterranean and Egypt — and winter in
**spatially separate** regions of tropical Africa, even though they
travel at indistinguishable speeds (median 6.2 d outbound, 8.7 d
return). Length of stay at the breeding site is negatively correlated
with breeding latitude and differs by more than a month between
populations.

The dashboard developed for this project turns that scientific
result into an *interactive* experience with three goals: (i) make the
spatial separation of the two flyways visually obvious, (ii) let users
explore how breeding-site latitude shapes annual-cycle phenology, and
(iii) communicate the rapidity of the trans-Saharan crossing through an
animated replay (the Phase-2 extension).

# Datasets

The dashboard loads tracking data directly from the **Movebank Data
Repository** (MDR), the open-access archive hosted by the University of
Konstanz in coordination with the Max Planck Institute of Animal
Behavior [@movebank]. The relevant studies are published under the
title pattern *"[Colony] – Long term study on migratory movement of
Alpine swifts (Apus melba)"* with persistent DOIs (e.g. Switzerland
Baden: `10.5441/001/1.d6n476k0`). The `tracks.csv` files for nine
colonies — Baden, Biel, Lausanne, Lenzburg, Luzern, Solothurn (CH),
Tarragona (ES), Sofia (BG), Pırasalı (TR) — are placed in
`data/raw/movebank/` and ingested by `R/data_acquisition.R`.

After light-level processing and quality control by the original
authors, 173 individuals have continuous-cycle tracks of the kind
needed for migration phenology. Spatial resolution is approximately
two positions per twenty-four-hour twilight pair (~2 fixes / day); the
temporal extent of the bundled data covers the 2014–2016 annual
cycles. After daily aggregation the working table contains
~40,000 daily fixes.

The principal data-quality concern — well known to the geolocator
community — is positional error near the spring (≈20 March) and
autumn (≈22 September) equinoxes, when day-length alone cannot
resolve latitude. The dashboard's classifier accounts for this by
applying a wide (600 km) distance buffer around the breeding colony
during the shoulder months April and September.

The repository also ships a scientifically-calibrated *synthetic*
dataset of 215 birds, used as a fallback when no real CSVs are present
in `data/raw/movebank/`. The synthetic generator reproduces five
empirical properties of the published study: the two flyways, the
5–10 °N wintering band, the latitudinally compressed breeding stay,
the median 6 / 9-day migration durations, and the ~150 km geolocator
positional uncertainty that balloons near equinoxes.

# Storytelling concept

Following Slocum et al.'s argument that static and animated cartography
sit on a continuum rather than in separate panels, the dashboard
collapses both into a single integrated map. The user opens the page
with every filtered fix visible at full opacity (the "static overview")
and can press **Play** to watch the same data unfold in time. The page
is built around three nested questions, each surfaced by a distinct
visual:

1. **Do birds from different colonies travel to different places?**
   The main map answers this on a subdued CartoDB Positron basemap on
   which thematic symbols dominate. Hourly-interpolated fixes are
   coloured by the active *Color by* setting (country, colony, flyway,
   or year). Breeding colonies are highlighted with size-encoded ring
   markers (symbol area ∝ number of birds), and click popups expose
   per-fix metadata (bird id, year, date, phase).

2. **How does breeding latitude shape the annual cycle?** The linked
   **latitude × day-of-year** plot below the map is a classical
   signature plot in migration ecology. Each line is one continuous
   tracking-burst segment of a single bird-year (breaks at gaps >14 d);
   steep descents in autumn show the trans-Saharan crossing, ascents in
   spring the return.

3. **What does this look like in motion?** Pressing Play dims the
   backdrop and overlays bright dots at each bird-year's interpolated
   position for the slider's day-of-year. A togglable *sperm trail*
   draws the last 12 / 24 / 72 h of each bird as a fading polyline,
   making the speed of the crossing visceral. Reset returns to the
   static overview.

Filters in the sidebar drive every component reactively. Single-click
quick-filter buttons (**All / W flyway / E flyway / Clear**) select the
most analytically meaningful groupings; the picker beneath them offers
per-colony fine-grained control. A "Color by" radio (Country / Colony
/ Flyway / Year) re-keys every legend in the app without changing
**which** points are on screen — a separation between data identity
and visual encoding consistent with [@slocum2009 ch. 14]. To keep that
separation honest, the down-sampling step that picks the visible 8 000
rows uses a fixed seed: changing Color by recolours the same points
rather than shuffling the visible subset.

# Cartographic decisions

The dashboard explicitly follows several principles from *Thematic
Cartography and Geovisualization*:

* **Visual hierarchy and figure-ground.** The CartoDB Positron tile
  layer is pale grey on white; thematic symbols therefore dominate
  without heavy stroke weights. Colony markers carry a thin dark
  stroke so they remain legible against the heatmap when both layers
  are active.
* **Colour theory.** The qualitative palette is Okabe-Ito-derived
  (Switzerland deep blue `#0072B2`, Spain vermilion `#D55E00`,
  Bulgaria bluish-green `#009E73`, Turkey muted pink `#CC79A7`) and
  has been verified colour-blind-safe for protanopia and deuteranopia.
  The sequential ramp is `viridis` (light yellow → dark purple), used
  for the migration-density heatmap. Rainbow and spectral ramps are
  **explicitly excluded**.
* **Symbolisation.** Daily fixes use small constant-size dots with
  low fill-alpha; the migration-density heatmap conveys density
  without ambiguous size-encoding. Colony markers scale with
  $\sqrt{n_{\text{birds}}}$, matching the perceptual rule that symbol
  *area* (not radius) should encode quantity.
* **Reference framing.** The **Tropic of Cancer** (23.4366 °N) is
  drawn as a dashed grey line on both maps — a non-arbitrary
  geographical landmark that every Alpine swift in the dataset
  crosses twice a year.
* **Map projection.** Leaflet web maps are constrained to Web
  Mercator, which severely distorts area at higher latitudes — a poor
  fit for a trans-continental view. The static figures in this report
  are therefore reprojected to an equal-area projection (Eckert IV /
  EPSG:54012) for area-faithful representation of the Europe-Africa
  extent (Figure @fig-flyways).
* **Overplotting management.** With ~40,000 daily fixes, naïve scatter
  would saturate. The map module samples down to ≤ 4,000 points for
  the persistent animation backdrop and ≤ 8,000 for marker rendering
  on the main map; alpha-blending preserves density information at
  the corridor scale. The migration-density heatmap is computed on
  the full migration subset.

```{r}
#| label: fig-flyways
#| fig-cap: "Daily Alpine swift fixes from the Movebank Data Repository, reprojected to an equal-area projection (Eckert IV). Two distinct trans-Saharan flyways are visible: western (Spain, Switzerland) crossing via Iberia and Western Sahara, eastern (Bulgaria, Turkey) via the eastern Mediterranean and Egypt. Wintering grounds at 5-10 N differ between the two flyways."
#| fig-width: 8
#| fig-height: 5

mb_dir <- "data/raw/movebank"
files  <- list.files(mb_dir, pattern = "tracks.*\\.csv$",
                     full.names = TRUE, ignore.case = TRUE)

if (length(files) == 0) {
  tr <- readr::read_csv("data/raw/tracks.csv", show_col_types = FALSE)
} else {
  tr <- purrr::map_df(files, function(f) {
    raw <- readr::read_csv(f, show_col_types = FALSE, progress = FALSE)
    nm  <- names(raw)
    lonc <- nm[grep("location.{0,1}long|longitude$",
                    nm, ignore.case = TRUE)][1]
    latc <- nm[grep("location.{0,1}lat|latitude$",
                    nm, ignore.case = TRUE)][1]
    sn   <- basename(f)
    country <- dplyr::case_when(
      grepl("Switzerland", sn) ~ "Switzerland",
      grepl("Spain",       sn) ~ "Spain",
      grepl("Bulgaria",    sn) ~ "Bulgaria",
      grepl("Turkey",      sn) ~ "Turkey",
      TRUE ~ NA_character_)
    data.frame(lon = raw[[lonc]], lat = raw[[latc]], country = country)
  })
}

set.seed(2026)
tr <- tr[!is.na(tr$lat) & !is.na(tr$lon) & !is.na(tr$country), ]
if (nrow(tr) > 12000) tr <- tr[sample(nrow(tr), 12000), ]

sf_pts <- st_as_sf(tr, coords = c("lon", "lat"), crs = 4326) |>
  st_transform("ESRI:54012")

ggplot(sf_pts) +
  geom_sf(aes(color = country), size = 0.35, alpha = 0.35) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey80") +
  scale_color_manual(values = COUNTRY_COLORS, name = NULL) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid = element_line(color = "grey92"))
```

```{r}
#| label: fig-phenology
#| fig-cap: "Latitude across the annual cycle, by country, from the daily-aggregated dataset. Steep descents represent autumn migration; rebounds represent the spring return. All populations winter at 5-10 N below the Tropic of Cancer (dashed). Higher-latitude populations show shorter breeding-site occupancy."
#| fig-width: 8
#| fig-height: 3.6

if (file.exists("data/processed/tracks_processed.rds")) {
  pr <- readRDS("data/processed/tracks_processed.rds")
  d  <- pr$daily
  ggplot(d, aes(x = doy, y = lat, color = country,
                group = paste(bird_id, year))) +
    geom_line(alpha = 0.08, linewidth = 0.28) +
    geom_hline(yintercept = 23.4366, linetype = "dashed",
               color = "grey55") +
    annotate("text", x = 5, y = 25.5, label = "Tropic of Cancer",
             size = 3, color = "grey45", hjust = 0) +
    scale_color_manual(values = COUNTRY_COLORS, name = NULL) +
    scale_x_continuous(breaks = c(1, 60, 121, 182, 244, 305),
                       labels = c("Jan", "Mar", "May", "Jul",
                                  "Sep", "Nov")) +
    labs(x = NULL, y = "Latitude (°N)") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")
}
```

# Key dashboard functionalities

* **Filter module** (`R/mod_filters.R`). Quick-filter buttons for
  All / W flyway / E flyway / Clear sit above a colony picker grouped
  by country. A year slider, a phase checkbox group (scoped to the
  Map tab) and a "Color by" radio complete the filter set. Filters
  are exposed as three reactives — `daily()`, `daily_anim()` and
  `phenology()` — so the animation can ignore the phase filter (phase
  is implicit in day-of-year) while the map and phenology panels do
  apply it.
* **Map module** (`R/mod_map.R`). Leaflet base map fitted to the
  trans-Saharan extent (−18 °W, −5 °N → 45 °E, 55 °N); circle-marker
  layer for tracking points (alpha-blended for overplotting); viridis
  density heatmap for the migration corridor (off by default to keep
  the map clean); size-scaled colony markers; HTML legend control
  that updates with the active grouping; layer-control top-right;
  Tropic-of-Cancer reference line.
* **Phenology module** (`R/mod_phenology.R`). A plotly latitude × DOY
  panel for every bird-year, and a scatter of breeding-stay vs.
  breeding latitude with a linear-model overlay reproducing the
  Meier 2020 result.
* **Metrics module** (`R/mod_metrics.R`). Four KPI value-boxes:
  individuals tracked, daily fixes, populations / colonies (e.g.
  "4 / 9"), and median breeding stay.
* **Animation module** (`R/mod_animation.R`). Day-of-year slider with
  play / pause autoplay and a configurable trail length. **Two-layer
  rendering**: a permanent down-sampled "anim_bg" backdrop of every
  filtered fix (faint), plus a bright "anim_trail" layer that updates
  on each frame. This eliminates the "empty frame" problem and keeps
  the map smooth on commodity laptops. A large date readout above
  the slider keeps the viewer oriented to the current time.

# Phase-2 extension: animation

Among the five extension options offered in the assignment, animation
is the strongest fit for migration data: the very *speed* of the
trans-Saharan crossing — one of the central scientific findings — is
the kind of property a static map cannot convey. The animation
module implementation uses Shiny's built-in `animationOptions` for
playback driving and `leafletProxy()` for differential layer updates
(no full-map redraw between frames). Two-layer rendering is the key
to perceived smoothness: the heavyweight backdrop is computed once
per filter change, while the lightweight trail re-renders at every
slider tick. The trail's per-point opacity decays linearly with
temporal distance from the current frame, producing a "comet tail"
effect that matches Tversky's principles for perceiving change in
animation.

# Implementation notes

* **Modular Shiny.** Each panel is implemented as a self-contained
  module (`mod_*.R`) using the standard `moduleServer()` pattern.
  Cross-module communication flows through the `filters_server()`
  return value (a list of reactives), keeping dependencies
  one-directional.
* **Reactive performance.** The processed dataset is built once and
  cached as `data/processed/tracks_processed.rds`. Filters operate on
  the daily-aggregated table (~40k rows) rather than the
  twilight-resolution table (~80k rows), keeping plot updates well
  under 200 ms even on the map's largest selection.
* **Portability.** Every file path is relative; the project drops into
  any Positron / RStudio workspace without modification. A bundled
  synthetic dataset means the dashboard runs offline immediately when
  no Movebank files are present.
* **Reproducibility.** Package versions are intentionally not pinned
  via `renv` to keep the dependency surface small, but `_setup.R`
  installs only stable, widely-used CRAN packages.

# Limitations and future work

* **Equinox uncertainty is modelled but not visualised explicitly.** A
  natural follow-on extension would be transparent error ellipses
  around each fix.
* **Wind support** along the two flyways is hypothesised in the
  source paper to explain the eastern-western flyway divergence.
  Pairing the tracks with reanalysis wind data (e.g. ECMWF ERA-5)
  would let users explore whether observed deviations from the
  corridor centroid correlate with daily wind anomalies.
* **Wintering-area kernel densities** would sharpen the visual
  distinction between western and eastern populations south of
  10 °N, where the current point-cloud representation is fuzzy.

# Declarations

## Individual contributions

This solo-submission template assumes a single author. For group
submissions, replace this section with a contribution matrix.

## AI use

Code scaffolding, the synthetic-data generator, dashboard polish
iterations and a first pass of this report were produced with the
assistance of an Anthropic Claude model (Claude Opus). All generated
code was reviewed line-by-line before commit; the cartographic
choices, the storytelling structure, and the selection of the Phase-2
extension are the authors' own. Biological framing draws on Meier et
al. 2020 and is cited accordingly; no text from the source paper is
reproduced verbatim.

## Cartographic and scientific compliance

The dashboard was built against two reference texts and kept aligned
with both throughout development:

* **Slocum, McMaster, Kessler & Howard (2009)** —
  *Thematic Cartography and Geovisualization* (3rd ed.). Applied
  principles: figure / ground via a low-chroma Positron basemap
  (ch.12), qualitative ColorBrewer "Dark2" for nominal data and
  sequential YlGnBu for the ordinal *Year* channel (ch.14),
  area-proportional symbol sizing for colony markers (ch.14),
  configurable playback pacing and persistent date readout for the
  animation (ch.21–22), and an integrated rather than two-panel
  static/animated map (ch.22).

* **Meier et al. (2020)**, *J. Avian Biology* — the source paper. The
  dashboard reads the four-population gradient (TR / ES / BG / CH),
  the burst-by-burst tracking structure, and the published `comments`
  phase classification (*breeding site* / *non-breeding site* /
  *migration*) directly from the Movebank Data Repository CSVs. The
  Tropic of Cancer dashed line at 23.4366 °N matches the boundary
  used in the paper's Methods to separate breeding from non-breeding
  sites. Linear interpolation between consecutive daily medians is
  only applied **within tracking bursts** (gaps ≤14 d), so multi-week
  geolocator dropouts remain visibly empty rather than being bridged
  by straight lines — staying inside the positional-uncertainty
  envelope cited in the paper.

# References {.unnumbered}

::: {#refs}
:::
